---
title: "NAFLD Figures"
author: "Andrew Perry"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Import data

```{r}
library(tidyverse)
library(readr)
library(ComplexHeatmap)

cardia.rdata.file <- "/Users/perry/Library/CloudStorage/OneDrive-VUMC/Research/CARDIA/NAFLD/Coding Files/CARDIA_NAFLD.Rdata"

filesToMd5 <- c(cardia.rdata.file)

dataFileInfo=data.frame(
  File=basename(filesToMd5),
  md5=tools::md5sum(filesToMd5)
)

knitr::kable(dataFileInfo)
load(cardia.rdata.file)
```

# check if the HIV proteins made into the lasso coefs

```{r}
table(lasso.coefs %>% 
        left_join(protein.dictionary) %>%
        filter(HR3LAMEANageSexRaceBMI != 0) %>%
        pull(Organism)
)
```

Our LASSO model for HR3LAMEAN (continuous fatty liver measure) with forced adjustments for age/sex/race/bmi only contains HUMAN proteins.

# SteatoSITE output

```{r}
steato.site <- HR3LAMEAN_marty$ageSexRaceBMI


top50absBeta <- steato.site %>%
                  filter(val_fdr<0.05) %>%
                  mutate(coef.rank = rank(-abs(der_beta))) %>%
                  filter(coef.rank<=50) %>% pull(AptName)

top25posNegBeta <- steato.site %>% 
                  filter(val_fdr<0.05) %>%
                  mutate(coef.rank = rank(der_beta)) %>%
                  filter(coef.rank<=25 | coef.rank >=(length(coef.rank)-24)) %>% pull(AptName)

steato.site <- steato.site %>% mutate(allFDRsig = case_when(val_fdr<0.05 ~ 1,
                                                            val_fdr>=0.05 ~ 0),
                                      top50absBeta = case_when(AptName %in% top50absBeta ~ 1,
                                                               !AptName %in% top50absBeta ~ 0),
                                      top25posNegBeta = case_when(AptName %in% top25posNegBeta ~ 1,
                                                               !AptName %in% top25posNegBeta ~ 0)
)

readr::write_csv(steato.site,
                 "dataForSteatoSITE.csv")

tools::md5sum("dataForSteatoSITE.csv")
```

# Distribution of covariates, gghistogram by NAFLD status of the HR3LAMEAN

```{r}
temp <- dataForModels %>% select(!any_of(proteins.to.use))
sort(colnames(temp))
```

## Age
```{r}
ggplot(temp, aes(EX8_AGE, HR3LAMEAN)) +
  geom_point(aes(col=densCols(EX8_AGE, HR3LAMEAN, colramp = colorRampPalette(rev(rainbow(10, end = 4/6))))), position = "jitter") +
  scale_color_identity() +
  theme_bw() +
  ggpubr::stat_cor(cor.coef.name="rho", method="spearman") +
  labs(x="Age",
       y="Liver attenuation")
ggsave("HR3LAMEAN_vs_age_CARDIA.pdf", height = 4, width = 6, device = "pdf")
```


```{r}
ggplot(temp %>% mutate(NAFLD = factor(nafld, levels=0:1, labels=c("No", "Yes"))), aes(x=EX8_AGE)) +
  geom_histogram(aes(fill=NAFLD, col=NAFLD), position = "identity", alpha=0.25, binwidth = 1) +
  theme_bw() +
  labs(x="Age",
       y="Count",
       subtitle = "CARDIA Year 25") 
ggsave("NAFLD_vs_age_CARDIA.pdf", height = 4, width = 6, device = "pdf")
```

## BMI

```{r}
ggplot(temp, aes(exp(H20BMI), HR3LAMEAN)) +
  geom_point(aes(col=densCols(exp(H20BMI), HR3LAMEAN, colramp = colorRampPalette(rev(rainbow(10, end = 4/6)))))) +
  scale_color_identity() +
  theme_bw() +
  ggpubr::stat_cor(cor.coef.name="rho", method="spearman", label.x.npc = "right", hjust=1) +
  labs(x="BMI",
       y="Liver attenuation")
ggsave("HR3LAMEAN_vs_BMI_CARDIA.pdf", height = 4, width = 6, device = "pdf")

```

```{r}
ggplot(temp %>% mutate(NAFLD = factor(nafld, levels=0:1, labels=c("No", "Yes"))), aes(x=exp(H20BMI))) +
  geom_histogram(aes(fill=NAFLD, col=NAFLD), position = "identity", alpha=0.25) +
  theme_bw() +
  labs(x="BMI",
       y="Count",
       subtitle = "CARDIA Year 25")
ggsave("NAFLD_vs_age_CARDIA.pdf", height = 4, width = 6, device = "pdf")

```

## Sex and race distribution of HR3LAMEAN

```{r}
colorset <- c("Female" = "darkorange", "Male" = "forestgreen")

summary(lm(HR3LAMEAN ~ SEX*RACE, data=temp))

ggplot(temp, aes(x=HR3LAMEAN)) +
  geom_density(aes(colour = SEX, linetype = RACE)) +
  geom_vline(xintercept = 40, color="purple", linetype="dashed") +
  scale_fill_manual(values=colorset) +
  scale_colour_manual(values = colorset) +
  labs(x = "Liver attenuation\n(<40 ~ MASLD)",
       y = "Density",
       subtitle = "All CARDIA participants") +
  theme_bw() +
  guides(color=guide_legend(title = "Gender"),
         linetype=guide_legend(title = "Race"))
ggsave("HR3LAMEAN_vs_SexRace_CARDIA.pdf", height = 4, width = 6, device = "pdf")

```



# Plot volcano for the main model in pub format - use those that pass derivation and test in the validation set

```{r}
ggplot(HR3LAMEAN_marty$ageSexRaceBMI %>% filter(!is.na(val_fdr)), aes(x=val_beta, y=-log(val_p, base=10), label = EntrezGeneSymbol)) +
  geom_point(aes(col=val_fdr<0.05)) +
  ggsci::scale_color_npg(name = "FDR", labels = c(">=5%", "<5%")) +
  theme_bw() +
  ggrepel::geom_text_repel(size=2) +
  # guides(color=guide_legend(title="FDR<5%")) +
  # scale_color_discrete(name = "FDR", labels = c("≥5%", "<5%")) +
  labs(subtitle = "CARDIA Validation, N: 803",
       x=expression(CARDIA~Validation~beta),
       y=expression(-log[10](p)),
       caption = "Adjusted for age, gender, race, and BMI.\nFiltered to show proteins with FDR<5% in derivation.") +
  theme(legend.position = "right")
ggsave("nafld_validation_volcano.pdf")
```

Caption: Volcano plot of proteins related to liver attenuation (negative beta coefficients ~ more fatty liver). Proteins visualized passed FDR<5% in the derivation sample. Model results from the validation sample are used for visualization.


# Plot marty on marty for derive/validate

```{r}
temp <- HR3LAMEAN_marty$ageSexRaceBMI %>% mutate(col = densCols(HR3LAMEAN_marty$ageSexRaceBMI$der_beta, HR3LAMEAN_marty$ageSexRaceBMI$val_beta, colramp = colorRampPalette(rev(rainbow(10, end = 4/6)))))

cor(temp$der_beta, temp$val_beta, method = "spearman")
cor.test(temp$der_beta, temp$val_beta, method = "spearman")$p.value

ggplot(temp, aes(x=der_beta, y=val_beta)) +
  geom_point(aes(col=col)) +
  scale_color_identity() +
  ggpubr::stat_cor(method = "spearman", cor.coef.name = "rho") +
  theme_bw() +
  labs(x=expression(CARDIA~Derivation~beta),
       y=expression(CARDIA~Validation~beta))

ggsave("CARDIA_der_val_beta_on_beta.pdf")
rm(temp)
```

# Plot marty on marty for validation: main model vs. full model beta; identify how many continue to retain significance after full adjustment

```{r}
temp <- HR3LAMEAN_marty$ageSexRaceBMI %>% 
        select(where(is.character), contains("val")) %>% 
        rename_with(~gsub("val", "ageSexRaceBMI_val", .x), contains("val"))

temp2 <- HR3LAMEAN_marty$full %>% 
         select(where(is.character), contains("val")) %>%
         rename_with(~gsub("val", "full_val", .x), contains("val"))

temp <- temp %>% left_join(temp2) %>% mutate(col = densCols(temp$ageSexRaceBMI_val_beta, temp$full_val_beta, colramp = colorRampPalette(rev(rainbow(10, end = 4/6)))))

cor(temp$ageSexRaceBMI_val_beta, temp$full_val_beta, method = "spearman")
cor.test(temp$ageSexRaceBMI_val_beta, temp$full_val_beta, method = "spearman")$p.value

ggplot(temp, aes(ageSexRaceBMI_val_beta, full_val_beta)) +
  geom_point(aes(col=col)) +
  scale_color_identity() +
  theme_bw() +
  labs(x=expression(CARDIA~Validation~Age/Gender/Race/BMI~beta),
       y=expression(CARDIA~Validation~Full~beta)) +
  ggpubr::stat_cor(cor.coef.name = "rho", method="spearman")
ggsave("ageSexRaceBMI_full_beta_on_beta.pdf")

table(temp$ageSexRaceBMI_val_fdr<0.05, temp$full_val_fdr<0.05, exclude = NULL)
```

So we have 88+157+14=259 proteins that pass FDR<5% in the main (age/sex/race/bmi) model

And we have 160 that pass FDR<5% in the full model

# Heatmap of LASSO proteins and cluster
Annotation bar of +/- NAFLD, HRLAMEAN and BMI.  
I am using the age, sex, race, and BMI adjusted LASSO models.  

Since we have ~300 proteins in the lasso models for HR3LAMEAN, for visualization purposes I will include the top 50 (ranked by abs(beta)) in the heatmap

I am also going to filter the participants to show it in the validation sample.

```{r}
set.seed(12)
lasso.proteins <- lasso.coefs %>% 
                  filter(grepl("seq.", AptName)) %>% 
                  filter(coef.rank<=50) %>% pull(AptName)

marty.proteins <- HR3LAMEAN_marty$ageSexRaceBMI %>%
                  filter(val_fdr<0.05) %>%
                  mutate(coef.rank = rank(der_beta)) %>%
                  filter(coef.rank<=25 | coef.rank >(length(der_beta)-24)) %>% pull(AptName)

plotData <- dataForModels %>%
            filter(derive=="validation") %>%
            select(all_of(c("short_id", "H20BMI", "HR3LAMEAN", "nafld", marty.proteins))) %>%
            arrange(desc(HR3LAMEAN))

plotData$nafld <- factor(plotData$nafld,
                         levels = 0:1,
                         labels = c("No", "Yes"))

ha <- HeatmapAnnotation(`BMI` = exp(plotData$H20BMI),
                        `MASLD` = plotData$nafld,
                        `CT Liver attenutation` = plotData$HR3LAMEAN,
                        which = "column",
                        simple_anno_size = unit(0.3, "cm"),
                        annotation_name_gp = grid::gpar(fontsize = 6))


mat <- plotData %>% select(all_of(marty.proteins))
name.change <- data.frame(AptName = colnames(mat))
name.change <- name.change %>% left_join(protein.dictionary %>% select(AptName, EntrezGeneSymbol), by="AptName")

colnames(mat)[match(name.change$AptName, colnames(mat))] <- name.change$EntrezGeneSymbol
colnames(mat) <- name.change$EntrezGeneSymbol

pdf("NAFLD_heatmap_MartyProteins.pdf", height = 7, width = 11)
draw(
  Heatmap(as.matrix(t(mat)),
          name = "Normalized, log-transformed protein abundance",
          top_annotation = ha,
          cluster_columns = FALSE,
          clustering_method_rows = "ward.D2",
          # clustering_distance_rows = "ward.D2",
          # clustering_method_columns = "ward.D2",
          # clustering_distance_columns = "ward.D2",
          cluster_rows = TRUE,
          show_row_dend = FALSE,
          show_column_dend = FALSE,
          row_names_gp = grid::gpar(fontsize = 6),
          heatmap_legend_param = list(direction = "horizontal",
                                      title_position = "topcenter")),
  annotation_legend_side = "right",
  heatmap_legend_side = "bottom"
)
dev.off()
```

Caption: Proteins were selected by filtering to the proteins with FDR <5% in CARDIA derivation and validation then ranking on the derivation sample beta coef (top 50 abs(beta)). Columns are participants, ordered by liver attenutation. Rows are clustered (cluster method = "complete" distance = "euclidean"). Heatmap fill is the relative protein abundance (log transformed) in the validation sample.

# Show LASSO calibration across derivation and validation

top 21
```{r}
temp <- bind_rows(lasso.scores$HR3LAMEAN$ageSexRaceBMItop21, .id = "derive") %>% 
        left_join(dataForModels %>% select(short_id, HR3LAMEAN, nafld), by="short_id")

temp$nafld <- factor(temp$nafld,
                     levels = 0:1,
                     labels = c("No", "Yes"))
temp$derive <- factor(temp$derive,
                      levels = c("derive", "validate"),
                      labels = c("CARDIA Derivation, N: 1876", "CARDIA Validation, N: 803"))

ggplot(temp, aes(x=s1, y=HR3LAMEAN)) +
  geom_point(aes(color = densCols(s1, HR3LAMEAN, colramp = colorRampPalette(rev(rainbow(10, end = 4/6)))))) +
  scale_color_identity() +
  facet_wrap(.~derive) +
  geom_hline(yintercept = 40, color="black", linetype = "dashed") +
  theme_bw() +
  theme(aspect.ratio = 1) +
  labs(x = "Top 21 protein score of liver attenuation",
       y = "CT liver attenuation (HU)",
       caption = "Dashed line is the clincally used cutoff (<40) for MASLD") +
  ggpubr::stat_cor(method = "spearman",
                   cor.coef.name = "rho")
ggsave("lasso_calibration_top21_CARDIA.pdf", height = 4, width = 6, device = "pdf")
```

full
```{r}
temp <- bind_rows(lasso.scores$HR3LAMEAN$ageSexRaceBMI, .id = "derive") %>% 
        left_join(dataForModels %>% select(short_id, HR3LAMEAN, nafld), by="short_id")

temp$nafld <- factor(temp$nafld,
                     levels = 0:1,
                     labels = c("No", "Yes"))
temp$derive <- factor(temp$derive,
                      levels = c("derive", "validate"),
                      labels = c("CARDIA Derivation, N: 1876", "CARDIA Validation, N: 803"))

ggplot(temp, aes(x=s1, y=HR3LAMEAN)) +
  geom_point(aes(color = densCols(s1, HR3LAMEAN, colramp = colorRampPalette(rev(rainbow(10, end = 4/6)))))) +
  scale_color_identity() +
  facet_wrap(.~derive) +
  geom_hline(yintercept = 40, color="black", linetype = "dashed") +
  theme_bw() +
  theme(aspect.ratio = 1) +
  labs(x = "Protein score of liver attenuation",
       y = "CT liver attenuation (HU)",
       caption = "Dashed line is the clincally used cutoff (<40) for MASLD") +
  ggpubr::stat_cor(method = "spearman",
                   cor.coef.name = "rho")
ggsave("lasso_calibration_CARDIA.pdf", height = 4, width = 6, device = "pdf")

# ggplot(temp, aes(x=s1, y = ..count.., color = nafld, fill = nafld)) +
#   geom_histogram(alpha=0.25, position = "identity") +
#   facet_wrap(.~derive) +
#   ggsci::scale_fill_npg() +
#   ggsci::scale_color_npg() +
#   theme_bw() +
#   theme(legend.position = "bottom") +
#   labs(x = "Lasso score of liver attenuation",
#      y = "Density",
#      title = paste("CARDIA Derivation, N:", nrow(temp))) +
#   guides(color = guide_legend(title = "NAFLD"),
#          fill = guide_legend(title = "NAFLD"))

t.test(temp$s1[temp$derive=="CARDIA Derivation, N: 1876" & temp$nafld=="No"],
       temp$s1[temp$derive=="CARDIA Derivation, N: 1876" & temp$nafld=="Yes"])

t.test(temp$s1[temp$derive=="CARDIA Validation, N: 803" & temp$nafld=="No"],
       temp$s1[temp$derive=="CARDIA Validation, N: 803" & temp$nafld=="Yes"])

summary(lm(s1 ~ derive*nafld, data=temp))

ggplot(temp, aes(x=s1, y = ..density.., color = nafld, fill = nafld)) +
  geom_density(alpha=0.25, position = "identity") +
  facet_wrap(.~derive) +
  ggsci::scale_fill_npg() +
  ggsci::scale_color_npg() +
  theme_bw() +
  theme(legend.position = "bottom",
        aspect.ratio = 1) +
  labs(x = "Protein score of liver attenuation",
     y = "Density") +
  guides(color = guide_legend(title = "MASLD"),
         fill = guide_legend(title = "MASLD")) +
  annotate("text", x = -2, y = 1,
           label=expression(italic(p)<2.2e-16))
ggsave("lasso_by_NAFLD_CARDIA.pdf", height = 4, width = 6, device = "pdf")


```

# Show LASSO score by age, sex, race, BMI

```{r}
temp <- bind_rows(lasso.scores$HR3LAMEAN$ageSexRaceBMI, .id = "derive")
temp <- temp %>% left_join(dataForModels %>% select(all_of(c("short_id", adjustmentList$ageSexRaceBMIdrinks))), by="short_id")

# scale score
temp$s1 <- scale(temp$s1)
```


## AGE
```{r}
ggplot(temp, aes(y=s1, x=EX8_AGE)) +
  geom_point(aes(color = densCols(temp$EX8_AGE, temp$s1, colramp = colorRampPalette(rev(rainbow(10, end = 4/6))))), position = "jitter") +
  scale_color_identity() +
  theme_bw() +
  labs(y = "Protein score of liver attenuation",
       x = "Age",
       subtitle = paste("CARDIA Derivation & Validation, N:", nrow(temp))) +
  ggpubr::stat_cor(method = "spearman",
                   cor.coef.name = "rho") +
  theme(legend.position = "bottom")
ggsave("ageSexRaceBMI_lasso_HR3LAMEAN_age_DeriveValidate.pdf", height = 4, width = 6, device = "pdf")
```

## SEX & RACE
```{r}
colorset <- c("Female" = "darkorange", "Male" = "forestgreen")

summary(lm(s1 ~ SEX*RACE, data=temp))

ggplot(temp, aes(x=s1)) +
  geom_density(aes(colour = SEX, linetype = RACE)) +
  scale_fill_manual(values=colorset) +
  scale_colour_manual(values = colorset) +
  labs(x = "Protein score of liver attenuation",
       y = "Density",
       subtitle = paste("CARDIA Derivation & Validation, N:", nrow(temp))) +
  theme_bw() +
  guides(color=guide_legend(title = "Gender"),
         linetype=guide_legend(title = "Race")) +
  annotate("text", x = -4, y = 0.5, hjust = 0, vjust=1,
           label=expression(Gender~italic(p)==0.095)) +
  annotate("text", x = -4, y = 0.45, hjust=0, vjust=1,
           label=expression(Race~italic(p)==3.92e-7))
ggsave("ageSexRaceBMI_lasso_HR3LAMEAN_vs_SexRace_CARDIA.pdf", height = 4, width = 6, device = "pdf")
```


## BMI
```{r}
ggplot(temp, aes(y=s1, x=exp(H20BMI))) +
  geom_point(aes(color = densCols(exp(temp$H20BMI), temp$s1, colramp = colorRampPalette(rev(rainbow(10, end = 4/6)))))) +
  scale_color_identity() +
  theme_bw() +
  labs(y = "Protein score of liver attenuation",
       x = "BMI",
       subtitle = paste("CARDIA Derivation & Validation, N:", nrow(temp))) +
  ggpubr::stat_cor(method = "spearman",
                   cor.coef.name = "rho",
                   label.x.npc = "right",
                   hjust = 1)  +
  guides(color = guide_legend(title = "CARDIA Sample")) +
  theme(legend.position = "bottom")
ggsave("ageSexRaceBMI_lasso_HR3LAMEAN_BMI_DeriveValidate.pdf", height = 4, width = 6, device = "pdf")
```

## Alcohol use

```{r}
ggplot(temp %>% drop_na(drink_wk), aes(y=s1, x=drink_wk)) +
  geom_point(aes(color = densCols(temp$drink_wk, temp$s1, colramp = colorRampPalette(rev(rainbow(10, end = 4/6))))), position = "jitter") +
  scale_color_identity() +
  theme_bw() +
  labs(y = "Protein score of liver attenuation",
       x = "Alcoholic drinks per week",
       subtitle = paste("CARDIA Derivation & Validation, N:", nrow(temp %>% drop_na(drink_wk)))) +
  ggpubr::stat_cor(method = "spearman",
                   cor.coef.name = "rho",
                   label.x.npc = "right",
                   hjust = 1)  +
  guides(color = guide_legend(title = "CARDIA Sample")) +
  theme(legend.position = "bottom")
ggsave("ageSexRaceBMI_lasso_HR3LAMEAN_etohdrinks_DeriveValidate.pdf", height = 4, width = 6, device = "pdf")
```


# Show how the score is related to liver attenutation across categories of BMI

```{r}
temp <- temp %>% 
        filter(derive=="validate") %>%
        mutate(Category = case_when(exp(H20BMI) <30 ~ "Normal/Overweight",
                                    exp(H20BMI) >=30 ~ "Obese")) %>%
        left_join(cardia.data %>% select(short_id, HR3LAMEAN, HR3VISFATVOL))

table(temp$Category, exclude = NULL)

ggplot(temp, aes(s1, HR3LAMEAN, color=Category)) +
  geom_point(size=0.7) +
  theme_bw() +
  ggsci::scale_color_npg() +
  ggpubr::stat_cor(method = "spearman", cor.coef.name = "rho", show.legend = FALSE) +
  geom_hline(yintercept = 40, color="black", linetype="dashed") +
  labs(y="CT liver attenuation (HU)",
       x="Protein score of liver attenuation",
       subtitle = paste0("CARDIA Validation, N: ", nrow(temp),
                         "\n  Normal/Overweight, N: ", nrow(temp %>% filter(Category=="Normal/Overweight")),
                         "\n  Obese, N: ", nrow(temp %>% filter(Category=="Obese"))),
       caption = "Dashed line is the clincally used cutoff (<40) for MASLD") +
  guides(color=guide_legend(title = "")) +
  theme(legend.position = "bottom")
ggsave("lasso_calibration_by_ov-ob.pdf", height = 4, width = 6, device = "pdf")


ggplot(temp, aes(HR3VISFATVOL, s1)) +
  geom_point()
```

```{r}
summary(lm(HR3LAMEAN ~ s1, data=temp))

summary(lm(HR3LAMEAN ~ s1+H20BMI, data=temp))

summary(lm(HR3LAMEAN ~ s1*H20BMI, data=temp))
```

# ROC Curves

I am trying to make something prettier than this

```{r, eval=FALSE}
plot(model.1_roc, col = "red");
  lines(model.2_roc, col = "blue");
  lines(model.3_roc, col="green");
  legend("bottomright", legend = c("Model 1 (clinical)", "Model 2 (full protein score)", "Model 3 (top 21 protein score"),
       col = c("red", "blue", "green"), lty = 1)
```

```{r, eval=TRUE}
mod1 <- data.frame(Model = "Model 1 (clinical)",
                   Sensitivity = model.1_roc$sensitivities,
                   Specificity = model.1_roc$specificities)

mod2 <- data.frame(Model = "Model 2 (full protein score)",
                   Sensitivity = model.2_roc$sensitivities,
                   Specificity = model.2_roc$specificities)

mod3 <- data.frame(Model = "Model 3 (top 21 protein score)",
                   Sensitivity = model.3_roc$sensitivities,
                   Specificity = model.3_roc$specificities)

temp <- bind_rows(mod1, mod2, mod3)

ggplot(temp, aes(Specificity, Sensitivity)) +
  geom_line(aes(col=Model)) +
  ggsci::scale_color_npg() +
  xlim(1, 0) +
  theme_bw() +
  theme(aspect.ratio = 1,
        legend.position = "bottom") +
  guides(color = guide_legend(title = "")) +
  annotate("text", x = 0.75, y = 0.5,
           label = "AUC:\nModel 1 (clinical): 0.84 (0.80-0.88)\nModel 2 (full protein score): 0.94 (0.92-0.96)\nModel 3 (top 21 protein score): 0.94 (0.93-0.96)", hjust=0, vjust=1, size=2.5) +
  labs(subtitle = "CARDIA Validation, N: 793")
ggsave("ROC_CARDIA_Validate.pdf", device = "pdf")
```

Base clinical model: age/sex/race/BMI
Model 1: age/sex/race/BMI/drinks + AST + ALT + scale(log(HbA1c))
Model 2: age/sex/race/BMI/drinks + full protein score
Model 3: age/sex/race/BMI/drnink + top21 protein score

```{r, eval=FALSE}
library(pROC)
roc.list <- roc(nafld ~ model.1.class.prob + model.2.class.prob + model.3.class.prob, data = validation.data)


ci.list <- lapply(roc.list, ci.se, specificities = seq(0, 1, l = 25))

dat.ci.list <- lapply(ci.list, function(ciobj) 
  data.frame(x = as.numeric(rownames(ciobj)),
             lower = ciobj[, 1],
             upper = ciobj[, 3]))

names(roc.list) <- c("Model 1 (clinical)", "Model 2 (full protein score)", "Model 3 (top 21 protein score)")

p <- ggroc(roc.list) + 
  theme_bw() + 
  ggsci::scale_fill_npg() +
  geom_abline(slope=1, intercept = 1, linetype = "dashed", alpha=0.7, color = "grey") + 
  coord_equal()

for(i in 1:3) {
  p <- p + geom_ribbon(
    data = dat.ci.list[[i]],
    aes(x = x, ymin = lower, ymax = upper),
    fill = i + 1,
    alpha = 0.2,
    inherit.aes = F) 
  } 

p +
  annotate("text", x = 0.75, y = 0.25,
           label = "AUC:\nModel 1 (clinical): 0.84\nModel 2 (full protein score): 0.94\nModel 3 (top 21 protein score): 0.94", hjust=0, vjust=1, size=3) +
  theme(legend.position = "bottom") +
  guides(color = guide_legend(title = "")) +
  labs(subtitle = "CARDIA Validation, N: 793",
       y="Sensitivity",
       x="Specificity")
ggsave("ROC_CARDIA_Validate.pdf", device = "pdf")
```

# Save output

```{r}
save.image("NAFLD_Figures.Rdata")
tools::md5sum("NAFLD_Figures.Rdata")

sessionInfo()
```