---
title: "NAFLD"
author: "Andrew Perry"
date: '`r Sys.Date()`'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Import data

```{r}
library(tidyverse)
library(gtsummary)
library(caret)
library(doParallel)
library(foreach)
library(nricens)
library(pROC)

somaDataFile <- "/Users/perry/Library/CloudStorage/OneDrive-VUMC/Research/CARDIA/Proteomics QC Files/CARDIA_Y25_Final_Proteomics.Rdata"
somaDictionaryFile <- "/Users/perry/Library/CloudStorage/OneDrive-VUMC/Research/CARDIA/Proteomics QC Files/CARDIA_Y25_Proteomics_DictionaryQC.csv"
CARDIAFile <- "/Users/perry/Library/CloudStorage/OneDrive-VUMC/Research/CARDIA/Data from Laura/CARDexpoV5.csv"
ukbbOverlapDictionary <- "/Users/perry/Library/CloudStorage/OneDrive-VUMC/Research/CARDIA/Exercise proteomics/Coding Files/CARDIA_UKBB_overlapDictionary.csv"

load(somaDataFile)
protein.dictionary <- readr::read_csv(somaDictionaryFile)
cardia.data <- readr::read_csv(CARDIAFile)
ukbb.overlaps <- readr::read_csv(ukbbOverlapDictionary)

filesToMd5 <- c(somaDataFile,
                somaDictionaryFile,
                CARDIAFile)

dataFileInfo=data.frame(
  File=basename(filesToMd5),
  md5=tools::md5sum(filesToMd5)
)

knitr::kable(dataFileInfo)
outputFilesToMd5 <- NULL
```

# Add missing EntrezGeneSymbols to the protein dictionary

I manually looked these up at https://www.genenames.org

```{r}
protein.dictionary$EntrezGeneSymbol <- case_when(protein.dictionary$EntrezGeneID=="645142" ~ "PPIAL4D",
                                                 protein.dictionary$EntrezGeneID=="101180976" ~ "IFNL4",
                                                 protein.dictionary$EntrezGeneID=="100128071" ~ "FAM229A",
                                                 protein.dictionary$EntrezGeneID=="401285" ~ "TCP10L2",
                                                 protein.dictionary$EntrezGeneID=="100131244" ~ "ANKRD63",
                                                 protein.dictionary$EntrezGeneID=="147646" ~ "C19orf84",
                                                 protein.dictionary$EntrezGeneID=="648791" ~ "PPP1R3G",
                                                 protein.dictionary$EntrezGeneID=="83715" ~ "ESPN",
                                                 protein.dictionary$EntrezGeneID=="8209" ~ NA,
                                                 protein.dictionary$EntrezGeneID=="100134938" ~"UPK3BL1" ,
                                                 protein.dictionary$EntrezGeneID=="2651" ~ NA,
                                                 protein.dictionary$EntrezGeneID=="402117" ~ "VWC2L",
                                                 protein.dictionary$EntrezGeneID=="140862" ~ "ISM1",
                                                 protein.dictionary$EntrezGeneID=="143678" ~ "FREY1",
                                                 protein.dictionary$EntrezGeneID=="647087" ~ "STMP1",
                                                 .default = protein.dictionary$EntrezGeneSymbol)
```


# Remove non-Human, non-proteins, high CV, the random CARDIA participant
```{r}
cv.cutoff=20

# I was advised to remove this participant from all studies
cardia.data <- cardia.data %>% filter(ID!=203323105195)

# find the non-human, non-protein stuff
table(protein.dictionary$Type)
nonProteins <- protein.dictionary %>% 
               filter(Type!="Protein") %>% 
               pull(AptName)

proteinsRemovedCV <- protein.dictionary %>% 
                     filter(interplate.cv > cv.cutoff | median.intraplate.cv > cv.cutoff) %>% 
                     pull(AptName)

dim(CARDIA_Y25_Final_Proteomics)
CARDIA_Y25_Final_Proteomics <- CARDIA_Y25_Final_Proteomics %>%
                               select(!any_of(c(nonProteins, proteinsRemovedCV)))
dim(CARDIA_Y25_Final_Proteomics)
```

I removed `r length(nonProteins)` non-human or non-proteins.

I removed `r length(proteinsRemovedCV)` proteins for CV > `r cv.cutoff`%.

log and scale proteins
```{r}
proteins.to.use <- colnames(CARDIA_Y25_Final_Proteomics %>% select(starts_with("seq")))
CARDIA_Y25_Final_Proteomics[,proteins.to.use] <- scale(log(CARDIA_Y25_Final_Proteomics[,proteins.to.use]))
```


# Exclude participants with other liver disease

**From Laua Colangelo (NW)**

  Attached is the dataset CARDexpoV4. This is CARDexpoV3 with the new variables added.
   
  Below are a few definitions and notes:
   
  drink_wk = sum(h07wine,h07beer,h07liqr);
  if h07drink=1 then drink_wk=0;
   
  if sex=1 then heavy_drink=( drink_wk > 21);
  if sex=2 then heavy_drink=( drink_wk > 14);
   
  ex_med25=(amiodarone25=1 or valproic25=1 or
            methotrex25=1 or tamox25=1 or diltia25=1);
   
  anyHEPabc = (h08hep=2 or h08hepa=2 or h08hepb=2 or h08hepc=2);
  onlyHEPc =(h08hepc=2);
   
  EXCLUDE1 =(ex_med25=1 or  heavy_drink=1 or anyHEPabc=1 or h08liver=2 or h08cir=2 or h08hiv=2);
  EXCLUDE2 =(ex_med25=1 or  heavy_drink=1 or onlyHEPc=1 or h08cir=2 or h08hiv=2);
   
   
  I worked on a van Wagner paper that is earlier than the one you linked to.
  In it, we excluded for hepatitis C.  James Terry seems to be suggesting excluding all of hepatitis A, B, and C.
  So I created the two exclusion variables EXCLUDE1 and EXCLUDE2.
  
**Andrew's plan**

I will use the convention used in this paper: https://www.ahajournals.org/doi/full/10.1161/JAHA.119.014279

This corresponds to EXCLUDE2 variable

```{r}
# change ID formatting on proteomics dataset
CARDIA_Y25_Final_Proteomics <- CARDIA_Y25_Final_Proteomics %>% 
                               mutate(short_id = gsub("-", "", SubjectID)) %>%
                               relocate(short_id, .before = everything())
CARDIA_Y25_Final_Proteomics$short_id <- gsub("Y25","", CARDIA_Y25_Final_Proteomics$short_id)

cardia.data$short_id <- as.character(cardia.data$short_id)
# decode certain variables into factors
cardia.data$SEX <- factor(cardia.data$SEX, levels = c(1,2), labels = c("Male", "Female"))
cardia.data$RACE <- factor(cardia.data$RACE, levels = c(4,5), labels = c("Black", "White"))
cardia.data$diab8 <- factor(cardia.data$diab8, levels = c(0,1), labels = c("No", "Yes"))
cardia.data$bpmed8 <- factor(cardia.data$bpmed8, levels = c(0,1), labels = c("No", "Yes"))
cardia.data$cholmed8 <- factor(cardia.data$cholmed8, levels = c(0,1), labels = c("No", "Yes"))

# add in the center coding for the clinical data
# 1=BHAM, 2=CHIC, 3=MINN, 4=OAKL
cardia.data$CENTER <- factor(cardia.data$CENTER, levels = 1:4, labels = c("BHAM", "CHIC", "MINN", "OAKL"))
cardia.data$center25 <- factor(cardia.data$center25, levels = 1:4, labels = c("BHAM", "CHIC", "MINN", "OAKL"))

dataForModels <- cardia.data %>%
                 inner_join(CARDIA_Y25_Final_Proteomics, by="short_id")

dim(dataForModels)
dataForModels <- dataForModels %>%
                 filter(EXCLUDE2==0)
dim(dataForModels)
```

So we excluded 275 individuals due to other conditions that cause hepatic steatosis

Create binary variable for NAFLD

**from Greg Terry (VUMC)**

There should also be variable called HR3LAMEAN that is mean of all those ROIs. That is the key variable for continuous attenuation. Of course, <=40 HU is cutpoint used for NAFLD or  <51 is equivalent to liver/spleen ratio <1.0 and is often used too. 

**Andrew's plan**

We will use the cutoff of ≤40, which is in line with the VanWagner paper above

```{r}
table(is.na(dataForModels$HR3LAMEAN))
dim(dataForModels)
dataForModels <- dataForModels %>%
                 drop_na(HR3LAMEAN) %>% # exclude those without steatosis measurements
                 mutate(nafld = ifelse(HR3LAMEAN <=40, 1, 0))
dataForModels$nafld <- as.factor(dataForModels$nafld)
dim(dataForModels)
```

So we lost 11 participants for lacking liver fat measures. 

# Define adjustments

```{r}
adjustmentList <- list(unadjusted = NULL,
                       
                       ageSexRace = c("EX8_AGE",
                                      "SEX",
                                      "RACE"),
                       
                       ageSexRaceBMI = c("EX8_AGE",
                                          "SEX",
                                          "RACE",
                                          "H20BMI"),
                       
                       ageSexRaceBMIdrinks = c("EX8_AGE",
                                               "SEX",
                                               "RACE",
                                               "H20BMI",
                                               "drink_wk"),
                       full = c("EX8_AGE",
                                 "SEX",
                                 "RACE",
                                 "H20BMI",
                                 "drink_wk",
                                "H02CSAVG",
                                "bpmed8",
                                "cholmed8",
                                "diab8",
                                "ltsmk25",
                                "HL1CHOL",
                                "HL1HDL",
                                "HL7GFR",
                                "H18TOTAL")
                         )
```

# Check distributions

We will select complete cases based on our adjustments

```{r}
dim(dataForModels)
dataForModels <- dataForModels %>% drop_na(adjustmentList$ageSexRaceBMIdrinks)
dim(dataForModels)
```

```{r}
# pdf("nafld_distributions.pdf")
Hmisc::hist.data.frame(dataForModels[,c("HR3LAMEAN", adjustmentList$ageSexRaceBMIdrinks)])
# dev.off()
```

Why are there negative values for liver fat?  
Per Greg Terry via email on 7/21/23 "These folks have fatty, fatty liver. All good and expected!"

Log transform.  
BMI. 

After we print table 1 though

# Split Derive/Validate and Table 1
```{r}
table1vars <- c(adjustmentList$full, c("center25",
                                       "H02CDAVG",
                                        "H0YHBA1C",
                                        "CVHBHI25",
                                        "HR3LAMEAN",
                                        "nafld",
                                        "HR3VISFATVOL",
                                        "HR3SBCTFATVOL"))

label.key <- list(EX8_AGE ~ "Age",
                  SEX ~ "Sex",
                  RACE ~ "Race",
                  H20BMI ~ "BMI",
                  center25 ~ "CARIDA field center",
                  H02CSAVG ~ "Mean SBP",
                  H02CDAVG ~ "Mean DBP",
                  bpmed8 ~ "On HTN med.",
                  cholmed8 ~ "On lipid med.",
                  diab8 ~ "Diabetes",
                  ltsmk25 ~ "Lifetime pack-yrs smoking",
                  HL7GFR ~ "eGFR",
                  H0YHBA1C ~ "A1c",
                  HL1CHOL ~ "Total cholesterol",
                  HL1HDL ~ "HDL",
                  HR3LAMEAN ~ "Mean liver attenutation",
                  nafld ~ "NAFLD",
                  HR3VISFATVOL ~ "VAT",
                  HR3SBCTFATVOL ~ "SAT",
                  CVHBHI25 ~ "Sum of AHA LSS",
                  drink_wk ~ "Drinks/wk",
                  H18TOTAL ~ "Physical activity score (H18 total)")

# balance the derive/validate by liver fat
set.seed(1234)
trainIndex <- caret::createDataPartition(dataForModels$HR3LAMEAN, 
                                           p=0.70,
                                           list = FALSE,
                                           times = 1)
dataForModels$derive <- NA
dataForModels[trainIndex[,1], "derive"] <- "derivation"
dataForModels[-trainIndex[,1], "derive"] <- "validation"

gtsummary::tbl_summary(dataForModels,
                       include = all_of(table1vars),
                       label = label.key,
                       by="derive",
                       statistic = list(all_continuous() ~  "{median} ({p25}, {p75}); {p_miss}%", all_categorical() ~ "{n} ({p}%); {p_miss}%"),
                       missing = "no") %>%
                       add_overall() %>%
                       add_p()

dataForModels$H20BMI <- log(dataForModels$H20BMI)
```


# Marty Larsen function

```{r}
marty_larsen <- function(predictor, outcome, family, adjustments = NULL, input_data){
  
  ifelse(is.null(adjustments), rhs <- predictor ,rhs <- paste(predictor, paste(adjustments, collapse = "+"), sep = "+"))
  
  fmla <- as.formula(paste0(outcome,"~",rhs))
  temp <- glm(fmla, data=input_data, family = family)
  
  output <- data.frame(Outcome = outcome,
                       AptName = predictor,
                       UniProt = protein.dictionary$UniProt[protein.dictionary$AptName==predictor],
                       EntrezGeneSymbol = protein.dictionary$EntrezGeneSymbol[protein.dictionary$AptName==predictor],
                       Target = protein.dictionary$Target[protein.dictionary$AptName==predictor],
                       TargetFullName = protein.dictionary$TargetFullName[protein.dictionary$AptName==predictor],
                       beta = summary(temp)$coefficients[predictor,"Estimate"],
                       stdError = summary(temp)$coefficients[predictor,"Std. Error"],
                       p.val = summary(temp)$coefficients[predictor, grep("Pr(>|[[:lower:]]|)", colnames(summary(temp)$coefficients), value = TRUE)],
                       nobs = nobs(temp))
  
  rm(temp); rm(rhs); rm(fmla);
  return(output)
}
```

# Lasso function

```{r}
fit_control <- trainControl(method="repeatedcv",
                            number=10,
                            repeats=5,
                            returnData=FALSE, 
                            savePredictions = "all", 
                            allowParallel = TRUE) 

lasso <- function(outcome, predictors, family, forcedAdjustments=NULL, input_data){
  if(is.null(forcedAdjustments)){
    rhs <- paste(predictors, collapse = "+")
  } else {
      rhs <- paste(c(forcedAdjustments, predictors), collapse = "+")
      }

  fmla<-as.formula(paste(outcome, "~", rhs))
  
  set.seed(1234)
  lasso.output <- caret::train(fmla,
                              data=input_data,
                              penalty.factor = c(rep(0, length(forcedAdjustments)), rep(1, length(predictors))),
                              method="glmnet",
                              family=family,
                              tuneGrid=expand.grid(alpha=1,
                                                   lambda= 10**(seq(-5,5,by=0.2))),
                              trControl=fit_control)
  
  return(lasso.output)
}
```

We will do marty larsens on all of the proteins in CARDIA

We will do lassos on all of the proteins in CARDIA. In a separate markdown I will recalibrate the score for use in UK Biobank.

```{r}
# lists to store output
marty.output <- NULL
lasso.models <- NULL
lasso.scores <- NULL
rfeModels <- NULL
rfe.scores <- NULL

#setup parallel processing
cores <- detectCores()-2
registerDoParallel(cores=cores)

outcome <- "nafld"
i <- names(adjustmentList)[4]

# loop over adjustments we could force into the model
for(outcome in c("HR3LAMEAN", "nafld")){
  for(i in names(adjustmentList)){
    family <- case_when(outcome=="HR3LAMEAN" ~ "gaussian",
                        outcome=="nafld" ~ "binomial")
    
    derive <- dataForModels %>% 
              filter(derive == "derivation") %>%
              select(all_of(c("short_id", outcome, adjustmentList[[i]], proteins.to.use))) %>% 
              filter(complete.cases(.))
    
    validate <- dataForModels %>% 
              filter(derive == "validation") %>%
              select(all_of(c("short_id", outcome, adjustmentList[[i]], proteins.to.use))) %>% 
              filter(complete.cases(.))
    
    # scale derivation and validation separately. recall I had scaled the whole cohort above
    derive[,c(proteins.to.use)] <- scale(derive[,c(proteins.to.use)])
    validate[,c(proteins.to.use)] <- scale(validate[,c(proteins.to.use)])
    
    # only scale the continuous outcome
    if(outcome == "HR3LAMEAN"){
      derive[,c(outcome)] <- scale(derive[,c(outcome)])
      validate[,c(outcome)] <- scale(validate[,c(outcome)])
    }
    
    ##
    ## Marty Larsen
    ##
    # Marty's on the derivation sample
    marty.output[[outcome]][[i]][["derive"]] <- foreach(k=proteins.to.use, .combine = "rbind") %dopar% {
    marty_larsen(outcome=outcome, predictor=k, family = family, adjustments = adjustmentList[[i]], input_data=derive)
    }
    # Marty's on the validation sample
    marty.output[[outcome]][[i]][["validate"]] <- foreach(k=proteins.to.use, .combine = "rbind") %dopar% {
    marty_larsen(outcome=outcome, predictor=k, family = family, adjustments = adjustmentList[[i]], input_data=validate)
    }
    
    # don't need to do LASSO models on the "full" adjustment set
    if(i=="full"){next}
  
    ##
    ## LASSO
    ##
    # Create lasso models using the derivation sample
    lasso.models[[outcome]][[i]] <- lasso(outcome=outcome,
                                        predictors=proteins.to.use,
                                        family = family,
                                        forcedAdjustments=adjustmentList[[i]],
                                        input_data = derive %>% select(!short_id))
    
    # pull out the coefficients for the proteins
    temp <- data.frame(as.matrix(coef(lasso.models[[outcome]][[i]]$finalModel, lasso.models[[outcome]][[i]]$bestTune$lambda))) %>% rownames_to_column(var="AptName")
    coef.vector <- as.vector(temp$s1[grepl("seq", temp$AptName)])
    
    
    # calculate & store the derivation samples lasso scores
    
    # make sure things are lined up
    if(!identical(colnames(derive %>% select(proteins.to.use)), temp$AptName[grepl("seq", temp$AptName)])){
      print("ERROR: LASSO score construction is wrong. Matrix colnames do not match coef.vector")
    }
    
    lasso.scores[[outcome]][[i]][["derive"]] <- data.frame(short_id = derive$short_id,
                                           s1 = as.matrix(derive %>% select(all_of(proteins.to.use))) %*% coef.vector
    )
    
    
    # calculate & store the validation samples lasso scores
    
    # make sure things are lined up
    if(!identical(colnames(validate %>% select(proteins.to.use)), temp$AptName[grepl("seq", temp$AptName)])){
      print("ERROR: LASSO score construction is wrong. Matrix colnames do not match coef.vector")
    }
    
    lasso.scores[[outcome]][[i]][["validate"]] <- data.frame(short_id = validate$short_id,
                                             s1 = as.matrix(validate %>% select(all_of(proteins.to.use))) %*% coef.vector
    )
    
    ##
    ## RFE (removed on 8/14/23 due to skepticism in RFE results in other analyses)
    ##
    ## I am using this to create several RFE models (like creating several LASSO models above) to determine how to best 
    ## select a panel of 21 proteins. At present, I do not think it is possible to "force" variables into an RFE. 
    ## It would be really attractive (I think) to do RFE to select 21 proteins that are conditioned on a set of adjustments
    ## (e.g., age, sex, race, BMI) but I don't think that is possible. Alternatively, I will use RFE to select 21+ features
    ## and examine whether the adjustments are selected in the models.
    ##
    ## Recall that RFE simply creates an LM model after selecting the most important variables. So I could identify 21 proteins
    ## then put those into an lm model alongside the adjustments of choice to create my model/score.
    
    # create our model formula
    # if(is.null(adjustmentList[[i]])){
    #   rhs <- paste(proteins.to.use, collapse = "+")
    # } else {
    #     rhs <- paste(c(adjustmentList[[i]], proteins.to.use), collapse = "+")
    # }
    # fmla <- paste(outcome, "~", rhs)
    # # fmla <- as.formula(fmla)
    #   
    # if(outcome=="HR3LAMEAN"){
    #   ctrl <- rfeControl(functions = lmFuncs, # this needs to change based on outcome liver fat v nafld
    #                      method = "cv",
    #                      number=5,
    #                      verbose = FALSE,
    #                      allowParallel = TRUE)
    # } else {
    #   ctrl <- rfeControl(functions = lrFuncs, # this needs to change based on outcome liver fat v nafld
    #                      method = "cv",
    #                      number=5,
    #                      verbose = FALSE,
    #                      allowParallel = TRUE)
    # }
    # 
    # set.seed(1234)
    # rfeModels[[outcome]][[i]][["full"]] <- caret::rfe(as.formula(fmla),
    #                                               #     x = derive %>% select(all_of(c(adjustmentList[[i]], proteins.to.use))),
    #                                               # y = derive %>% pull(outcome),
    #                                            data = derive,
    #                                            sizes = 1:length(proteins.to.use), # RFE over all proteins
    #                                            rfeControl = ctrl)
    # 
    # # best size on minimizing RMSE
    # max.size <- pickSizeBest(rfeModels[[outcome]][[i]][["full"]]$results, metric="Accuracy", maximize = FALSE)
    # # Tolerance 5% of minimal RMSE
    # min.size <- pickSizeTolerance(rfeModels[[outcome]][[i]][["full"]]$results, metric="Accuracy", tol=5, maximize = FALSE)
    # 
    # #
    # # Now do 5% tolerated RFE
    # # 
    # set.seed(1234)
    # rfeModels[[outcome]][[i]][["tol5"]] <- rfe(as.formula(fmla),
    #                                            # x = derive %>% select(all_of(c(adjustmentList[[i]], proteins.to.use))),
    #                                            #    y = derive %>% pull(outcome),
    #                                            data = derive,
    #                                            sizes = 1:length(proteins.to.use),
    #                                            size = min.size,
    #                                            rfeControl = ctrl)
    # 
    # #
    # # Now do RFE for 21 proteins
    # # 
    # set.seed(1234)
    # rfeModels[[outcome]][[i]][["olink21"]] <- rfe(as.formula(fmla),
    #                                               # x = derive %>% select(all_of(c(adjustmentList[[i]], proteins.to.use))),
    #                                               # y = derive %>% pull(outcome),
    #                                            data = derive,
    #                                            sizes = 1:length(proteins.to.use),
    #                                            size = 21 + length(adjustmentList[[i]]),
    #                                            rfeControl = ctrl)
    # 
    # #
    # # calculate RFE scores in derive and validate
    # #
    # if(outcome == "HR3LAMEAN"){
    #   rfe.scores[[outcome]][[i]][["derive"]] = data.frame(short_id = derive$short_id,
    #                                                       rfe.full.score = predict(rfeModels[[outcome]][[i]][["full"]], newdata=derive),
    #                                                       rfe.tol5.score = predict(rfeModels[[outcome]][[i]][["tol5"]], newdata=derive),
    #                                                       rfe.olink21.score = predict(rfeModels[[outcome]][[i]][["olink21"]], newdata=derive))
    #   
    #   rfe.scores[[outcome]][[i]][["validate"]] = data.frame(short_id = validate$short_id,
    #                                                         rfe.full.score = predict(rfeModels[[outcome]][[i]][["full"]], newdata=validate),
    #                                                         rfe.tol5.score = predict(rfeModels[[outcome]][[i]][["tol5"]], newdata=validate),
    #                                                         rfe.olink21.score = predict(rfeModels[[outcome]][[i]][["olink21"]], newdata=validate))
    # } else {
    #   rfe.scores[[outcome]][[i]][["derive"]] = data.frame(short_id = derive$short_id,
    #                                                       rfe.full.score = predict(rfeModels[[outcome]][[i]][["full"]], newdata=derive) %>% pull("1"),
    #                                                       rfe.tol5.score = predict(rfeModels[[outcome]][[i]][["tol5"]], newdata=derive) %>% pull("1"),
    #                                                       rfe.olink21.score = predict(rfeModels[[outcome]][[i]][["olink21"]], newdata=derive) %>% pull("1"))
    #   
    #   rfe.scores[[outcome]][[i]][["validate"]] = data.frame(short_id = validate$short_id,
    #                                                         rfe.full.score = predict(rfeModels[[outcome]][[i]][["full"]], newdata=validate) %>% pull("1"),
    #                                                         rfe.tol5.score = predict(rfeModels[[outcome]][[i]][["tol5"]], newdata=validate) %>% pull("1"),
    #                                                         rfe.olink21.score = predict(rfeModels[[outcome]][[i]][["olink21"]], newdata=validate) %>% pull("1"))
    # }
    
    rm(derive, validate, coef.vector, temp);
  }
}

stopImplicitCluster()
```

# Lambda check

Check that the lambda is not at the edge of the search grid

```{r}
lasso.coefs <- NULL

for(i in names(lasso.models)){
  for(j in names(lasso.models[[i]])){
    print(paste(i,j))
    print(range(lasso.models[[i]][[j]][["pred"]][["lambda"]]))
    print(range(lasso.models[[i]][[j]][["bestTune"]][["lambda"]]))
    
    temp <- data.frame(as.matrix(coef(lasso.models[[i]][[j]]$finalModel, lasso.models[[i]][[j]]$bestTune$lambda))) %>% rownames_to_column(var="AptName")
    colnames(temp)[colnames(temp)=="s1"] <-paste0(i,j)
    lasso.coefs[[paste0(i,"_",j)]] <- temp
    
    rm(temp)
  }
}

lasso.coefs <- plyr::join_all(lasso.coefs, type="full", by="AptName")
lasso.coefs <- lasso.coefs %>% 
               left_join(protein.dictionary %>% select(AptName, UniProt, EntrezGeneSymbol, Target, TargetFullName), by="AptName") %>% 
               relocate(UniProt, EntrezGeneSymbol, Target, TargetFullName, .after = AptName) %>%
               arrange(AptName)

scores <- colnames(lasso.coefs %>% select(contains("HR3LAMEAN") | contains("nafld")))
knitr::kable(apply(lasso.coefs %>% filter(grepl("seq.", AptName)) %>% select(all_of(scores)), 2, function(x) sum(x!=0, na.rm=TRUE)), caption = "Number of proteins in LASSO models")
```

The above supports the lambda search grid is appropriate. The optimal lambda is not near the edges.

# create a 21 protein score

```{r}
# identify the top 21 protein coefs from the age, sex, race, bmi adjusted lasso model
lasso.coefs <- lasso.coefs %>% 
                filter(!is.na(UniProt)) %>% 
                mutate(coef.rank = rank(-abs(HR3LAMEANageSexRaceBMI))) %>% 
                mutate(HR3LAMEANageSexRaceBMItop21 = case_when(coef.rank<=21 ~ HR3LAMEANageSexRaceBMI,
                                                               .default = 0))

derive <- dataForModels %>% 
          filter(derive == "derivation") %>%
          select(all_of(c("short_id", outcome, adjustmentList$ageSexRaceBMI, proteins.to.use))) %>% 
          filter(complete.cases(.))

validate <- dataForModels %>% 
          filter(derive == "validation") %>%
          select(all_of(c("short_id", outcome, adjustmentList$ageSexRaceBMI, proteins.to.use))) %>% 
          filter(complete.cases(.))

# scale derivation and validation separately. recall I had scaled the whole cohort above
derive[,c(proteins.to.use)] <- scale(derive[,c(proteins.to.use)])
validate[,c(proteins.to.use)] <- scale(validate[,c(proteins.to.use)])

# top21 score in derivation
# make sure things are lined up
if(!identical(colnames(derive %>% select(lasso.coefs$AptName[grepl("seq", lasso.coefs$AptName)])), lasso.coefs$AptName[grepl("seq", lasso.coefs$AptName)])){
  print("ERROR: LASSO score construction is wrong. Matrix colnames do not match coef.vector")
}
coef.vector <- lasso.coefs$HR3LAMEANageSexRaceBMItop21[grepl("seq", lasso.coefs$AptName)]

lasso.scores[["HR3LAMEAN"]][["ageSexRaceBMItop21"]][["derive"]] <- data.frame(short_id = derive$short_id,
                                         s1 = as.matrix(derive %>% select(lasso.coefs$AptName[grepl("seq", lasso.coefs$AptName)])) %*% coef.vector
)

# top 21 score in validation                                         
# make sure things are lined up
if(!identical(colnames(validate %>% select(lasso.coefs$AptName[grepl("seq", lasso.coefs$AptName)])), lasso.coefs$AptName[grepl("seq", lasso.coefs$AptName)])){
  print("ERROR: LASSO score construction is wrong. Matrix colnames do not match coef.vector")
}
coef.vector <- lasso.coefs$HR3LAMEANageSexRaceBMItop21[grepl("seq", lasso.coefs$AptName)]

lasso.scores[["HR3LAMEAN"]][["ageSexRaceBMItop21"]][["validate"]] <- data.frame(short_id = validate$short_id,
                                         s1 = as.matrix(validate %>% select(lasso.coefs$AptName[grepl("seq", lasso.coefs$AptName)])) %*% coef.vector
)
```


# Compare Marty's and LASSO coefs

```{r}
temp <- marty.output$HR3LAMEAN$ageSexRaceBMI$derive %>%
        left_join(lasso.coefs %>% select(where(is.character), HR3LAMEANageSexRaceBMI))

ggplot(temp, aes(HR3LAMEANageSexRaceBMI, beta)) +
  geom_point() +
  theme_bw() +
  labs(x="LASSO coef",
       y="Marty coef",
       subtitle = "derivation")

temp <- marty.output$HR3LAMEAN$ageSexRaceBMI$validate %>%
        left_join(lasso.coefs %>% select(where(is.character), HR3LAMEANageSexRaceBMI))

ggplot(temp, aes(HR3LAMEANageSexRaceBMI, beta)) +
  geom_point() +
  theme_bw() +
  labs(x="LASSO coef",
       y="Marty coef",
       subtitle = "validation")

rm(temp)
```

# Model fit

Plot scatters of scores against liver fat in derivation and validation, report R/rho

```{r}
for(i in names(lasso.scores)){
  for(j in names(lasso.scores[[i]])){
    for(k in names(lasso.scores[[i]][[j]])){
      # merge in the data
      temp <- lasso.scores[[i]][[j]][[k]] %>% left_join(dataForModels %>% select(all_of(c("short_id", i))), by="short_id")
      # scale score
      temp$s1 <- scale(temp$s1)
      
      if(i == "HR3LAMEAN"){
        print(
          ggplot(temp, aes(x=s1, y=HR3LAMEAN)) +
            geom_point(size=0.3) +
            theme_classic() +
            labs(x = paste(j, "score of", i, "in CARDIA", k),
                 y = i,
                 title = paste("N:", nrow(temp))) +
            ggpubr::stat_cor(method = "spearman",
                             cor.coef.name = "rho") +
            stat_smooth(formula = y ~ x,
                        method = "lm",
                        se = FALSE,
                        linetype = "dashed")
        )
      }
        
      if(i == "nafld"){
        print(
          ggplot(temp, aes(x=s1, y = ..count.., color = nafld, fill = nafld)) +
            geom_histogram(alpha=0.25, position = "identity") +
            theme_classic() +
            labs(x = paste(j, "score of", i, "in CARDIA", k),
               y = "Count",
               title = paste("N:", nrow(temp)))
        )
      }
        
      
      # rm(temp)
    }
  }
}
```

Plot relations of scores against age, sex, race, BMI

```{r}
for(i in names(lasso.scores)){
  for(j in names(lasso.scores[[i]])){
    temp <- bind_rows(lasso.scores[[i]][[j]], .id = "derive")
    temp <- temp %>% left_join(dataForModels %>% select(all_of(c("short_id", adjustmentList$ageSexRaceBMI))), by="short_id")
    
    # scale score
    temp$s1 <- scale(temp$s1)
    
    # AGE
    print(
      ggplot(temp, aes(y=s1, x=EX8_AGE, color = derive)) +
        geom_point(position = "jitter", size=0.05) +
        theme_classic() +
        labs(y = paste(j, "score of", i, "in CARDIA, derive+validate"),
             x = "Age",
             title = paste("N:", nrow(temp))) +
        ggpubr::stat_cor(method = "spearman",
                         cor.coef.name = "rho") +
        stat_smooth(formula = y ~ x,
                    method = "lm",
                    se = FALSE,
                    linetype = "dashed")
    )
    
    # SEX
    print(
      ggplot(temp, aes(x=s1, y = ..count.., color = SEX, fill = SEX)) +
        geom_histogram(alpha=0.25, position = "identity") +
        theme_classic() +
        labs(x = paste(j, "score of", i, "in CARDIA, derive+validate"),
           y = "Count",
           title = paste("N:", nrow(temp)))
    )
    # RACE
    print(
      ggplot(temp, aes(x=s1, y = ..count.., color = RACE, fill = RACE)) +
        geom_histogram(alpha=0.25, position = "identity") +
        theme_classic() +
        labs(x = paste(j, "score of", i, "in CARDIA, derive+validate"),
           y = "Count",
           title = paste("N:", nrow(temp)))
    )
    
    # BMI
    print(
      ggplot(temp, aes(y=s1, x=H20BMI, color = derive)) +
        geom_point(position = "jitter", size=0.05) +
        theme_classic() +
        labs(y = paste(j, "score of", i, "in CARDIA, derive+validate"),
             x = expression(log(BMI)),
             title = paste("N:", nrow(temp))) +
        ggpubr::stat_cor(method = "spearman",
                         cor.coef.name = "rho") +
        stat_smooth(formula = y ~ x,
                    method = "lm",
                    se = FALSE,
                    linetype = "dashed")
    )
  }
}
```

# Validation odds of liver fat on top of the overall model

What are some published prediction models for NAFLD?

https://www.ncbi.nlm.nih.gov/pmc/articles/PMC9381583/

https://www.sciencedirect.com/science/article/pii/S0168827817324364?via%3Dihub

https://www.nature.com/articles/s41591-019-0665-2

My take away from this prior literature is that a score using age, sex, race, bmi, AST, ALT, A1c would be a very reasonable clinical model to compare our proteomics against.

**Prediction of NAFLD in holdout set using the LASSO**
Specify logistic model - 
**Base clinical model:** age/sex/race/BMI  
**Model 1:** age/sex/race/BMI/drinks + AST + ALT + scale(log(HbA1c))  
**Model 2:** age/sex/race/BMI/drinks + AST + ALT + scale(log(HbA1c)) + full protein score    
**Model 3:** age/sex/race/BMI/drnink + AST + ALT + scale(log(HbA1c)) + top21 protein score   


Get the ORs and 95% CI
Get the confusion Matrix for the full model and 21 model; compare to the model 1 and model 2 (make a nice table, sens, spec, accuracy, etc.)

```{r}
validate <- dataForModels %>% 
              filter(derive == "validation") %>%
              select(all_of(c("short_id", outcome, adjustmentList$ageSexRaceBMIdrinks, "seq.4912.17", "seq.3709.4", "H0YHBA1C"))) %>% 
              left_join(lasso.scores$HR3LAMEAN$ageSexRaceBMI$validate, by="short_id") %>%
              rename(HR3LAMEANageSexRaceBMI = s1) %>%
              left_join(lasso.scores$HR3LAMEAN$ageSexRaceBMItop21$validate, by="short_id") %>%
              rename(HR3LAMEANageSexRaceBMItop21 = s1) %>%
              filter(complete.cases(.))

hist(validate$H0YHBA1C)
hist(scale(log(validate$H0YHBA1C)))

base.model = glm(nafld ~ EX8_AGE + SEX + RACE + scale(H20BMI), family = "binomial", data = validate, x=TRUE)
model.1 = glm(nafld ~ EX8_AGE + SEX + RACE + scale(H20BMI) + drink_wk + scale(seq.4912.17) + scale(seq.3709.4) + scale(log(H0YHBA1C)), family = "binomial", data = validate, x=TRUE)
model.2 = glm(nafld ~ EX8_AGE + SEX + RACE + scale(H20BMI) + drink_wk + scale(seq.4912.17) + scale(seq.3709.4) + scale(log(H0YHBA1C)) + scale(HR3LAMEANageSexRaceBMI), family = "binomial", data = validate, x=TRUE)
model.3 = glm(nafld ~ EX8_AGE + SEX + RACE + scale(H20BMI) + drink_wk + scale(seq.4912.17) + scale(seq.3709.4) + scale(log(H0YHBA1C)) + scale(HR3LAMEANageSexRaceBMItop21), family = "binomial", data = validate, x=TRUE)

# capture the model coefs, std error, z, p values
model.summaries <- as.data.frame(summary(base.model)$coefficients) %>% rownames_to_column(var="covariate") %>% mutate(model = "base.model")
model.summaries <- model.summaries %>% bind_rows(as.data.frame(summary(model.1)$coefficients) %>% rownames_to_column(var="covariate") %>% mutate(model = "model.1"))
model.summaries <- model.summaries %>% bind_rows(as.data.frame(summary(model.2)$coefficients) %>% rownames_to_column(var="covariate") %>% mutate(model = "model.2"))
model.summaries <- model.summaries %>% bind_rows(as.data.frame(summary(model.3)$coefficients) %>% rownames_to_column(var="covariate") %>% mutate(model = "model.3"))

# Prediction using the logistic models
# this gives a probability of having NAFLD
validation.data <- validate
validation.data$base.model.class.prob <- predict(base.model, newdata = validation.data, type = "response")
validation.data$model.1.class.prob <- predict(model.1, newdata = validation.data, type = "response")
validation.data$model.2.class.prob <- predict(model.2, newdata = validation.data, type = "response")
validation.data$model.3.class.prob <- predict(model.3, newdata = validation.data, type = "response")

# create factor variables for y/n prediction of NAFLD based on probability from above
prob.cutoff <- 0.50
validation.data$base.model.class <- case_when(validation.data$base.model.class.prob>=prob.cutoff ~ 1,
                                              validation.data$base.model.class.prob<prob.cutoff ~ 0)
validation.data$model.1.class <- case_when(validation.data$model.1.class.prob>=prob.cutoff ~ 1,
                                              validation.data$model.1.class.prob<prob.cutoff ~ 0)
validation.data$model.2.class <- case_when(validation.data$model.2.class.prob>=prob.cutoff ~ 1,
                                              validation.data$model.2.class.prob<prob.cutoff ~ 0)
validation.data$model.3.class <- case_when(validation.data$model.3.class.prob>=prob.cutoff ~ 1,
                                              validation.data$model.3.class.prob<prob.cutoff ~ 0)

validation.data$base.model.class <- as.factor(validation.data$base.model.class)
validation.data$model.1.class <- as.factor(validation.data$model.1.class)
validation.data$model.2.class <- as.factor(validation.data$model.2.class)
validation.data$model.3.class <- as.factor(validation.data$model.3.class)

output.matrix <- matrix(nrow = 4, ncol=18)
rownames(output.matrix) <- c("base.model", "model.1", "model.2", "model.3")
# confusion matrices
for(i in c("base.model", "model.1", "model.2", "model.3")){
  print(i)
  test <- caret::confusionMatrix(data = validation.data[[paste0(i,".class")]], reference = validation.data$nafld, positive="1")
  print(test)
  output.matrix[i,] <- unname(c(test$overall, test$byClass))
}
colnames(output.matrix) <- c(names(test$overall), names(test$byClass))

htmlTable::htmlTable(t(round(output.matrix, 2)))

output.matrix <- as.data.frame(output.matrix)
output.matrix <- output.matrix %>% rownames_to_column(var="model")

rmarkdown::paged_table(output.matrix)
```

Check the model calibration (Hosmer Lemeshow test)
Base model
```{r}
performance::performance_hosmer(base.model)
```

Model 1
```{r}
performance::performance_hosmer(model.1)
```

Model 2
```{r}
performance::performance_hosmer(model.2)
```

Model 3
```{r}
performance::performance_hosmer(model.3)
```

Get the NRI for model 3 vs. model 2 and model 1 (you can do two NRIs)

```{r}
NRI.Cstat.Output <- tibble(old.model = character(),
                           new.model = character(),
                           old.cindex = character(),
                           new.cindex = character(),
                           cindex.compare.pval = double(),
                           NRI.overall = character(),
                           NRI.events = character(),
                           NRI.nonevents = character())


nri.model.2.model.1 <- nribin(mdl.std = model.1,
                               mdl.new = model.2,
                               cut=0.50,
                              niter=5)

nri.summary <- paste0(signif(nri.model.2.model.1$nri$Estimate, 2), " (",signif(nri.model.2.model.1$nri$Lower, 2) , "-", signif(nri.model.2.model.1$nri$Upper, 2) , ")")

NRI.Cstat.Output <- NRI.Cstat.Output %>% add_row(old.model = "model.1",
                                                 new.model = "model.2",
                                                 NRI.overall = nri.summary[1],
                                                 NRI.events = nri.summary[2],
                                                 NRI.nonevents = nri.summary[3])

nri.model.3.model.1 <- nribin(mdl.std = model.1,
                               mdl.new = model.3,
                               cut=0.50,
                              niter=5)

nri.summary <- paste0(signif(nri.model.3.model.1$nri$Estimate, 2), " (",signif(nri.model.3.model.1$nri$Lower, 2) , "-", signif(nri.model.3.model.1$nri$Upper, 2) , ")")

NRI.Cstat.Output <- NRI.Cstat.Output %>% add_row(old.model = "model.1",
                                                 new.model = "model.3",
                                                 NRI.overall = nri.summary[1],
                                                 NRI.events = nri.summary[2],
                                                 NRI.nonevents = nri.summary[3])
```

Create ROC curve in publication format (model 1, model 2, model 3) or at least model 1 and Model 3;

```{r}
# ROC construction of base to full model

model.1_roc = plot(roc(validation.data$nafld ~ validation.data$model.1.class.prob, plot = TRUE, print.auc = TRUE))
model.2_roc = plot(roc(validation.data$nafld ~ validation.data$model.2.class.prob, plot = TRUE, print.auc = TRUE))
model.3_roc = plot(roc(validation.data$nafld ~ validation.data$model.3.class.prob, plot = TRUE, print.auc = TRUE))

# Plot the ROC curves on the same plot
plot(model.1_roc, col = "red"); lines(model.2_roc, col = "blue"); lines(model.3_roc, col="green"); legend("bottomright", legend = c("Model 1 (clinical)", "Model 2 (full protein score)", "Model 3 (top 21 protein score"),
       col = c("red", "blue", "green"), lty = 1)

NRI.Cstat.Output$old.cindex[NRI.Cstat.Output$new.model=="model.2"] <- paste0(round(model.1_roc$auc, 2), " (", round(as.numeric(ci(model.1_roc))[1], 2), "-", round(as.numeric(ci(model.1_roc))[3], 2), ")")
NRI.Cstat.Output$old.cindex[NRI.Cstat.Output$new.model=="model.3"] <- paste0(round(model.1_roc$auc, 2), " (", round(as.numeric(ci(model.1_roc))[1], 2), "-", round(as.numeric(ci(model.1_roc))[3], 2), ")")
NRI.Cstat.Output$new.cindex[NRI.Cstat.Output$new.model=="model.2"] <- paste0(round(model.2_roc$auc, 2), " (", round(as.numeric(ci(model.2_roc))[1], 2), "-", round(as.numeric(ci(model.2_roc))[3], 2), ")")
NRI.Cstat.Output$new.cindex[NRI.Cstat.Output$new.model=="model.3"] <- paste0(round(model.3_roc$auc, 2), " (", round(as.numeric(ci(model.3_roc))[1], 2), "-", round(as.numeric(ci(model.3_roc))[3], 2), ")")
NRI.Cstat.Output$cindex.compare.pval[NRI.Cstat.Output$new.model=="model.2"] <- roc.test(model.1_roc, model.2_roc)$p.value
NRI.Cstat.Output$cindex.compare.pval[NRI.Cstat.Output$new.model=="model.3"] <- roc.test(model.1_roc, model.3_roc)$p.value

roc.test(model.1_roc, model.2_roc)
roc.test(model.1_roc, model.3_roc)
roc.test(model.2_roc, model.3_roc)

rmarkdown::paged_table(NRI.Cstat.Output)
```

# Save output

## output Marty results

```{r}
HR3LAMEAN_marty <- NULL
nafld_marty <- NULL

for(i in names(marty.output$HR3LAMEAN)) {
  derive <- marty.output[["HR3LAMEAN"]][[i]][["derive"]] %>%
            rename(der_beta = beta,
                   der_stdError = stdError,
                   der_p = p.val,
                   der_n = nobs)
  validate <- marty.output[["HR3LAMEAN"]][[i]][["validate"]] %>%
              rename(val_beta = beta,
                     val_stdError = stdError,
                     val_p = p.val,
                     val_n = nobs)
  
  HR3LAMEAN_marty[[i]] <- derive %>% 
                          left_join(validate) %>% 
                          mutate(der_fdr = p.adjust(der_p, method="BH")) %>% 
                          group_by(der_fdr<0.05) %>% 
                          mutate(val_fdr = p.adjust(val_p)) %>% 
                          mutate(val_fdr = case_when(der_fdr>=0.05 ~ NA,
                                                    .default = val_fdr)) %>%
                          relocate(der_fdr, .after=der_p) %>%
                          relocate(val_fdr, .after=val_p) %>%
                          ungroup() %>%
                          select(!where(is.logical))
}

openxlsx::write.xlsx(HR3LAMEAN_marty, file = "HR3LAMEAN_Cardia_lm_protein.xlsx")

for(i in names(marty.output$nafld)) {
  derive <- marty.output[["nafld"]][[i]][["derive"]] %>%
            rename(der_beta = beta,
                   der_stdError = stdError,
                   der_p = p.val,
                   der_n = nobs)
  validate <- marty.output[["nafld"]][[i]][["validate"]] %>%
              rename(val_beta = beta,
                     val_stdError = stdError,
                     val_p = p.val,
                     val_n = nobs)
  
  nafld_marty[[i]] <- derive %>% 
                      left_join(validate) %>% 
                      mutate(der_fdr = p.adjust(der_p, method="BH")) %>% 
                      group_by(der_fdr<0.05) %>% 
                      mutate(val_fdr = p.adjust(val_p)) %>% 
                      mutate(val_fdr = case_when(der_fdr>=0.05 ~ NA,
                                                .default = val_fdr)) %>%
                      relocate(der_fdr, .after=der_p) %>%
                      relocate(val_fdr, .after=val_p) %>%
                      ungroup() %>%
                      select(!where(is.logical))
}

openxlsx::write.xlsx(nafld_marty, file = "nalfd_Cardia_lm_protein.xlsx")

outputFilesToMd5 <- c(outputFilesToMd5, "HR3LAMEAN_Cardia_lm_protein.xlsx", "nalfd_Cardia_lm_protein.xlsx")
```

## Output MD5 and Session Info
```{r}
outputFilesToMd5 <- c(outputFilesToMd5, "CARDIA_NAFLD.Rdata")
save.image("CARDIA_NAFLD.Rdata")

dataFileInfo=data.frame(
  File=basename(outputFilesToMd5),
  md5=tools::md5sum(outputFilesToMd5)
)

knitr::kable(dataFileInfo)


sessionInfo()
```
